TIM: Tree Imaging Machine for low-cost digital images of tree cores and cookies

### 2025-05-12 docs/writing/paper.tex

\usepackage{hyperref}          
\usepackage{subcaption}     
\usepackage{multirow}

@@

% Title section
\title{TIM: Tree Imaging Machine for low-cost digital images of tree cores and cookies} 
\author{Adam Fong,$^{1}$ Chloe Curry,$^{1}$ Sandy Zhang$^{1,2}$ and E.M. Wolkovich$^{1}$} %emw14Apr2025: You should ask Chloe and Sandy who they want their names displayed (middle initial etc.)
\date{\today}

\noindent $^{1}$Forest and Conservation Sciences, University of British Columbia, Vancouver, BC, Canada\\
$^{2}$Sandy's current affiliation\\ %emw14Apr2025 -- I am not sure you need to add this, usually the affiliation is where you did most of the work; it can be multiple if you use a significant # of employed at another job working on it (basically, the other university PAID your time to do this work)

@@

@@

\begin{abstract}  
Studies of tree rings from climatology to ecology have benefitted recently from new methods to image wood rapidly and efficiently, helping to scale up the number and quality of image samples. Current approaches are, however, limited in scanning dimensionality and can be cost-prohibitive. % Getting rid of open-source as CaptuRING is cheap but GigaPixel is expensive and not open source
To address these limitations, we designed the Tree Imaging Machine (TIM)---a do-it-yourself scanning tool to digitize tree cookies and cores. TIM takes partially overlapping microscopic images of samples and stitches the individual images together to form 
a mosaic, which can be zoomed in to visualize features on the scale of 0.01 mm. With scans of up to 21 140 DPI, TIM produces one of the highest resolution images among similar tools for less than \$3 000 USD. %emw14Apr2025 -- we're not good arbitrators of what is a competitive price 
We designed TIM to have a large working plane to allow digitizing cores over 50 cm and complete cookies. Operators can prepare sample batches of multiple cores and cookies, letting the machine work until the queue is finished.
%emw14Apr2025 -- add \href command below to make the URLs work 
All of the 3D printed parts are available for download on the [NIH 3D printing repository](https://doi.org/10.60705/3DPX/21561.3) and the software is  % is this best practice for including links? 
open source, with instructions to recreate this tool available on [GitHub](https://github.com/temporalecologylab/TreeRings). TIM provides an important advance in scanning resolution, tree cookie sampling, maximum core sample size, batch digitization, and cost in a single package. This promotes the digitization of larger samples and can collect data at finer resolutions in an open-source design which can be built on and improved as new technologies become available. % Archive GitHub with Zenodo? I'm nervous as the code base is updated relatively frequently... will this be a problem? Or do I only ever make one 'release' https://docs.github.com/en/repositories/archiving-a-github-repository/referencing-and-citing-content 
\end{abstract}

@@

Image analysis of tree rings was later implemented to reduce both errors in sampling and sampling bias across individual technicians, with major methods developed in the 1980s and 1990s to digitize tree ring samples using optical scanning \citep{mcmillin_application_1982} \citep{guay_new_1992}. Optical scanners are still readily available for purchase, using a top of the line scanner (e.g., Epson Perfection V850 Pro) produces a maximum resolutions of 6 400 dpi and scan area of up to 21.6 cm x 29.7 cm.
More recent techniques have aimed to achieve higher resolutions and/or larger samples through a different digitization approach---using an image-capturing system. These systems rely on a microscope camera to take images, motors to move the sample or camera, a computer to manage the capturing, and a second, more powerful, computer to stitch images into a larger composite image.
Decreasing the field of view of each captured image increases the resolution of the final stitched image without restricting the size of the sample \citep{muhlich_stitching_2022}.
Stitching multiple images into one mosaic is a common technique used in other fields such as mineralogy and cellular biology \citep{ro_image_2021,mohammadi_fast_2024}, which was
first implemented for tree rings through the Advanced Tree-Ring Image Capturing System (ATRICS) \citep{levanic_atrics_2007}. % Adding ATRICS after introducing the image and stitch method

@@

To address some of the current limitations, while building upon the open-source and open-hardware philosophies of CaptuRING with the functionality of Gigapixel, we designed the Tree Imaging Machine (TIM).
We added support to digitize both cookies and cores, increased the maximum sample length, perform image stitching directly without the need for manual file transfer, and capture batches of samples sequentially---all while minimizing cost.
Our design requires the use of 3D printed components and common hand tools such as Allen keys and wrenches. No specialized equipment or power tools are required beyond this. %emw14Apr2025 I think it's Allen wrenches and Hex keys --- check and update
Excluding 3D printed parts, the total cost of the machine is approximately \$2 200 USD \citep[compared to the \$70 000 USD of the Gigapixel,][]{griffin_gigapixel_2021}.
We chose to use a Raspberry Pi HQ camera with a microscope lens to save on purchasing a professional camera with a macro lens. Professional camera setups are comparable in price to the entire cost of TIM.
These savings combined with not needing to purchase a professional stitching software license (such as PTGui, the suggested software to use with CaptuRING) makes TIM a more economical option than most products current available and its open-source, open-hardware design makes it easy to adapt and adjust. %emw14Apr2025 -- add citation for PTGui? Check if their documentation suggests one....

@@

% Good topic sentence! Can you walk reader more carefully through each item? You may need to re-order them .. But basically, pick the most useful order and then write: For the camera .... For the gantry machine ... You might also use this space to say, in contrast to previous approaches that have used XX giant camera, we used ... and otherwise point out the novelty when it comes up. 
TIM can be thought of as a combination of multiple subsystems: a gantry machine, camera, and computer (Figure \ref{fig:tim_assembled}). 
For the gantry machine, cartesian movement in the $X$, $Y$ and $Z$ directions is a result of two machine kits and a motor controller from OpenBuilds---the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware. 
The machine kit includes extruded aluminum rails, carriages to slide along the rails, stepper motors to control the $X$ and $Y$ movement, and all the hardware to assemble the machine. 
To fit on the build plates of the ACRO system, we made an adapter to connect the linear actuator to the X and Y axis, thus creating motion in the Z axis.
The size of the work plane is ultimately up to the size of the ACRO kit that is purchased---allowing for more flexibility.
Building on top of kits allowed for quick assembly with sound instructions and saved development time. 
For the camera, we chose a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens and connected this to the gantry machine using custom adapters.
For the computer, we implemented all of the software control, image stitching, and graphical user interface (GUI) to run on an NVIDIA Jetson Orin Nano (Jetson), which also connects to the camera via a MIPI cable. %emw14Apr2025 -- I think we should avoid referring to products just by their brandname ... consider replacing Jetson with `Jetson edge computer' throughout. 
Controlling the machine is done by launching a Python program and interacting with mouse button clicks on the GUI.

@@

\begin{figure}
  \centering
  \begin{subfigure}{.3\textwidth}
      \centering
      \includegraphics[width=.95\linewidth]{../content/cookie_figure_grid.png}  
      \caption{}
      \label{SUBFIGURE LABEL 1}
  \end{subfigure}
  \begin{subfigure}{.3\textwidth}
      \centering
      \includegraphics[width=.95\linewidth]{../content/cookie_figure_traverse_start.png}  
      \caption{}
      \label{SUBFIGURE LABEL 2}
  \end{subfigure}
  \begin{subfigure}{.3\textwidth}
      \centering
      \includegraphics[width=.95\linewidth]{../content/cookie_figure_second_row.png}  
      \caption{}
      \label{SUBFIGURE LABEL 3}
  \end{subfigure} %emw14Apr2025: check my edits here please (capture region as a phrase okay?)!
  \caption{An example imaging process of a cookie sample. Here, the height and width of the capture region enclose the entire surface area of the cookie.  TIM begins by generating a grid of target coordinates to systematically capture a grid of continuous and partially overlapping images (a). TIM takes its first image at the center coordinate of the sample, then the rectangular field of view from the camera moves to adjacent target coordinates until it reaches the left boundary---capturing an image for each target coordinate (b). After completing the row, TIM moves to a new row and continues capturing adjacent images (c). }
  \label{FIGURE LABEL}
  \end{figure}

%emw14Apr2025: Careful with its (versus it's -- it is)!
The goal of this sample transverse is to obtain in-focus images that have a region of overlap with each of its neighbors---the basis of image stitching (Figure \ref{FIGURE LABEL}). 
The coordinates of each image-to-be-captured are derived from the sample's center coordinate, the height and width of the sample, the height and width of an image frame, and the user provided percentage overlap between images.  
Our design can capture cores by defining the sample width to be zero and the height of the core to be used. So long as the cores are aligned to be parallel to the Y axis, this minimizes imaging time as only one column of images is captured.
Digitizing cookies requires a rectangular grid of images to be taken in both the X and Y axes. 
The user provides enough information to find the X and Y coordinates of each image, but the Z coordinate must be found with an algorithmic focusing procedure.

@@

With TIM's microscope lens, the depth of field of the image is sensitive enough that sub-millimeter heights can move an image out-of-focus.
%emw14Apr2025: $Z_0$ is 'height' but in the notation table it's a vector ... I think maybe the table is wrong? 
This means that the height collected when adding a sample ($Z_0$, we provide a complete glossary of notation used to describe the methods, see Table \ref{tab:math_notation}) does not produce in-focus images throughout the entire surface area of the sample.
To capture an in-focus image at height $Z_{focused,k}$, for each unique image coordinate $(X_k, Y_k)$, we implemented a time-conscious software procedure. 
Our procedure involves two stages and starts at the known center coordinate of the sample.
The first stage takes advantage of the requirement that the sample is in-focus when adding it to be scanned.
Starting at the center coordinate allows the initial $Z_0$ value, captured when adding a sample, to find $Z_{focused,0}$. 
After finding an in-focus image, TIM navigates to the next neighboring image coordinate to find its in-focus image, using the previous neighbor's in-focus height as an initial approximation.

@@

The second stage of our time-efficient focusing procedure uses a feedback loop to update the initial $Z_k$ at each image location $(X_k, Y_k, Z_k)$. 
Without a feedback loop, the sample alignment must have a difference in height no greater than 0.5 mm ($d_1 - d_2 < 0.5 mm $ in Figure \ref{fig:realistic_levelling}). 
Rather than increasing this height range, a greater alignment error can be managed by adjusting the starting image height until $Z_{focused,k}$ is at the center index of $\boldsymbol{Z_{samples,k}}$. 
The likelihood of an adjacent image, at the coordinate $(X_{k+1}, Y_{k+1})$, containing an in-focus image is highest in this scenario. 
A negative feedback  %emw14Apr2025: check!
proportional-integral-derivative (PID_ control algorithm is used to adjust the initial height of each image, with $\alpha_k$, whenever the camera moves in the X or Y dimensions \citep{odwyer_summary_2000}. 
This alleviates the need of having remarkably level sample alignment ($d_1 - d_2 < 0.5\text{mm}$ in Figure \ref{fig:realistic_levelling}). %emw14Apr2025 -- check I did not mess up math notation here, if not fix above also so that mm is not italics
\[
Z_k = \begin{cases}
  Z_0 & \text{if } k = 0,\\
  Z_{k-1} + \alpha_k \\
  \end{cases}
\]
\[\boldsymbol{\delta} = 
\begin{bmatrix}
  0.5 \\
  0.4 \\
  \vdots \\
  -0.4 \\
  -0.5 \\
  \end{bmatrix}\]
\[
Z_{focused,k} \in
\boldsymbol{Z_k} = 
Z_k + \boldsymbol{\delta}
\]

@@

After testing numerous image stitching software APIs, we found the Python package Stitch2D most capable of stitching our images successfully. % need to reference this
This package wraps OpenCV functions for finding distinctive image features using the SIFT feature detector and matches these features within the overlapping region of adjacent images \citep{lowe_distinctive_2004}. 
The default implementation of the package produces high quality stitches but is memory inefficient. %is this wording correct?, I'm trying to kindly say it was memory inefficient
We implemented a few key memory-conscious changes to the package and were able to run it on the Jetson without a problem.

@@

%emw14Apr2025 -- this figure is not critical to the paper, so I would move it to supp and just reference it here (Fig S1 etc.)
\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[height=0.5\linewidth]{../diagrams/dpi_measurement_whole.png}
    \caption{Scale Bar for DPI measurements}
    \label{fig:dpi_measurement_whole}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[height=0.5\linewidth]{../diagrams/dpi_measurement_zoomed_2.png}
    \caption{Zoomed in Scale Bar}
    \label{fig:dpi_measurement_zoomed}
  \end{subfigure}
  \caption{Example of a single DPI measurement using a 0.01 mm slide scale. We found the DPI measured at multiple locations in the field of view vertically and horizontally  converged on the same value.}
  \label{fig:dpi_measurements}
\end{figure}

@@

\subsection{Strengths and Opportunities} %emw9Mar: And this section could become part of the Results & Discussion.... 
% add better opportunities
TIM's design minimizes the barriers to build an affordable, efficient and high-resolution image capture system in smaller labs without compromising scan quality, and while allowing scanning of cookies and cores. We have measured tree ring widths using the digitized samples from TIM using CooRecorder as well as ImageJ. %emw14Apr2025 cite ImageJ when first used in paper, see options here https://imagej.net/ij/docs/faqs.html
%emw9Mar: Any citations for developing repos to share tree ring images? Or just more on building better open shared data for tree rings? 
% It appears databases store data after images are processed and measured, potentially too expensive to store image datasets at high resolution?
Publicly available tree-ring datasets can be found on multiple online repositories such as TreeSource's Tree-Rings Database for Canadian sites or the International Tree-Ring Data Bank (ITRDB) for worldwide data \citep{girardin_national_2021} \citep{grissino-mayer_international_1997}.
Both of these databases offer metadata-rich data on measured tree-rings but do not store the source images of samples.
One opportunity for this development could come as more universal and large-scale open-data storage of tree ring images come online. 
Such open repositories would also increase the value of developing new frameworks count vessel or identify rings, which are already starting and have the potential to greatly decrease the time invested in manually identifying tree anatomy \citep{resente_mask_2021} \citep{polacek_automation_2023}. 
With these more fully developed TIM could be extended to be run such frameworks while scanning, possibly taking additional scans only when needed to reach sufficient algorithm accuracy. %emw9Mar -- check my riffing here.

% Sandy got back to me! :)
%emw14Apr2025  -- finalize the info below before sending outside the lab for review
In its current state, TIM only provides scanned images and requires users to import them into other tools for additional visual analyses, such as measuring tree-ring widths or counting vessels.
A useful addition to TIM would be to automatically detect tree-ring widths or other metrics while capturing images using computer vision models. We expect this may be possible as these methods improve in efficiency. Currently, we found that tree-ring detection by a recent machine learning algorithm  \citeauthor[R-CNN presented by][]{polacek_automation_2023} for a scanned cookie of size (HEIGHT)X(WIDTH) (WAITING FOR SANDY) pixels from TIM  took 15 minutes on a high-performance cluster (UBC ARC Sockeye) and 3.5 hours on a smaller server (RTX A6000 GPU server, which currently costs approximately \$3.50 USD for this). 
Although the Jetson is not as powerful as these servers, running these models in parallel to the image capturing algorithms would further reduce manual data annotation of tree-rings, and thus could help scale-up data collection.
% Further development would be required to run the deep learning models on the Jetson itself or with a web server to offload the heavier computation.

@@

%emw14Apr2025: Add acknowledgments here: Deirdre, Hoai Huong Nguyen Phan, Fredi, CRD, Mao etc.

\subsection{Supplemental Information}
\renewcommand{\thetable}{S\arabic{table}}

\include{notation_glossary_table}

>>>>

\usepackage{hyperref}          
\usepackage{subcaption}     
\usepackage{multirow}
\usepackage{float}

@@

% Title section
\title{TIM: Tree Imaging Machine for low-cost digital images of tree cores and cookies} 
\author{Adam Fong,$^{1}$ Chloe Curry,$^{1}$ Yue Zhang$^{1, 2}$ and E.M. Wolkovich$^{1}$} %emw14Apr2025: You should ask Chloe and Sandy who they want their names displayed (middle initial etc.)
\date{\today}

@@

\noindent $^{1}$Forest and Conservation Sciences, University of British Columbia, Vancouver, BC, Canada\\
$^{2}$Department of Computer Science and Engineering, University of California San Diego, La Jolla, CA, USA\\

@@

\begin{abstract}  
Studies of tree rings from climatology to ecology have benefitted recently from new methods to image wood rapidly and efficiently, helping to scale up the number and quality of image samples. Current approaches are, however, limited in scanning dimensionality and can be cost-prohibitive. % Getting rid of open-source as CaptuRING is cheap but GigaPixel is expensive and not open source
To address these limitations, we designed the Tree Imaging Machine (TIM)---a do-it-yourself scanning tool to digitize tree cookies and cores. TIM takes partially overlapping microscopic images of samples and stitches the individual images together to form 
a mosaic, which can be zoomed in to visualize features on the scale of 0.01 mm. With scans of up to 21 140 DPI, TIM produces one of the highest resolution images among similar tools for less than \$3 000 USD.
We designed TIM to have a large working plane to allow digitizing cores over 50 cm and complete cookies. Operators can prepare sample batches of multiple cores and cookies, letting the machine work until the queue is finished.

All of the 3D printed parts are available for download on the \href{https://doi.org/10.60705/3DPX/21561.3}{NIH 3D printing repository} and the software is  % is this best practice for including links? 
open source, with instructions to recreate this tool available on \href{https://github.com/temporalecologylab/TreeRings}{GitHub}. TIM provides an important advance in scanning resolution, tree cookie sampling, maximum core sample size, batch digitization, and cost in a single package. This promotes the digitization of larger samples and can collect data at finer resolutions in an open-source design which can be built on and improved as new technologies become available. % Archive GitHub with Zenodo? I'm nervous as the code base is updated relatively frequently... will this be a problem? Or do I only ever make one 'release' https://docs.github.com/en/repositories/archiving-a-github-repository/referencing-and-citing-content 
\end{abstract}

@@

Image analysis of tree rings was later implemented to reduce both errors in sampling and sampling bias across individual technicians, with major methods developed in the 1980s and 1990s to digitize tree ring samples using optical scanning \citep{mcmillin_application_1982,guay_new_1992}. Optical scanners are still readily available for purchase, using a top of the line scanner (e.g., Epson Perfection V850 Pro) produces a maximum resolutions of 6 400 dpi and scan area of up to 21.6 cm x 29.7 cm.
More recent techniques have aimed to achieve higher resolutions and/or larger samples through a different digitization approach---using an image-capturing system. These systems rely on a microscope camera to take images, motors to move the sample or camera, a computer to manage the capturing, and a second, more powerful, computer to stitch images into a larger composite image.
Decreasing the field of view of each captured image increases the resolution of the final stitched image without restricting the size of the sample \citep{muhlich_stitching_2022}.
Stitching multiple images into one mosaic is a common technique used in other fields such as mineralogy and cellular biology \citep{ro_image_2021,mohammadi_fast_2024}, which was
first implemented for tree rings through the Advanced Tree-Ring Image Capturing System (ATRICS) \citep{levanic_atrics_2007}. % Adding ATRICS after introducing the image and stitch method

@@

To address some of the current limitations, while building upon the open-source and open-hardware philosophies of CaptuRING with the functionality of Gigapixel, we designed the Tree Imaging Machine (TIM).
We added support to digitize both cookies and cores, increased the maximum sample length, perform image stitching directly without the need for manual file transfer, and capture batches of samples sequentially---all while minimizing cost.
Our design requires the use of 3D printed components and common hand tools such as hex keys and wrenches. No specialized equipment or power tools are required beyond this. 
Excluding 3D printed parts, the total cost of the machine is approximately \$2 200 USD \citep[compared to the \$70 000 USD of the Gigapixel,][]{griffin_gigapixel_2021}.
We chose to use a Raspberry Pi HQ camera with a microscope lens to save on purchasing a professional camera with a macro lens. Professional camera setups are comparable in price to the entire cost of TIM.
These savings combined with not needing to purchase a professional stitching software license (such as PTGui (New House Internet Services BV, Rotterdam, NL), the suggested software to use with CaptuRING) makes TIM a more economical option than most products current available and its open-source, open-hardware design makes it easy to adapt and adjust.

@@

% Good topic sentence! Can you walk reader more carefully through each item? You may need to re-order them .. But basically, pick the most useful order and then write: For the camera .... For the gantry machine ... You might also use this space to say, in contrast to previous approaches that have used XX giant camera, we used ... and otherwise point out the novelty when it comes up. 
TIM can be thought of as a combination of multiple subsystems: a gantry machine, camera, and computer (Figure \ref{fig:tim_assembled}). 
For the gantry machine, cartesian movement in the $X$, $Y$ and $Z$ directions is a result of two machine kits and a motor controller from OpenBuilds---the ACRO 1010, the NEMA 17 lead screw linear actuator, and the X32 Motor Controller running GRBL firmware. 
The machine kit includes extruded aluminum rails, carriages to slide along the rails, stepper motors to control the $X$ and $Y$ movement, and all the hardware to assemble the machine. 
To fit on the build plates of the ACRO system, we made an adapter to connect the linear actuator to the X and Y axis, thus creating motion in the Z axis.
The size of the work plane is ultimately up to the size of the ACRO kit that is purchased---allowing for more flexibility.
Building on top of kits allowed for quick assembly with sound instructions and saved development time. 
For the camera, we chose a 12MP Raspberry Pi HQ Camera equipped with a SEEED studio microscope lens and connected this to the gantry machine using custom adapters.
For the computer, we implemented all of the software control, image stitching, and graphical user interface (GUI) to run on an NVIDIA Jetson Orin Nano (Jetson edge computer), which also connects to the camera via a MIPI cable.  
Controlling the machine is done by launching a Python program and interacting with mouse button clicks on the GUI.

@@

\begin{figure}
  \centering
  \begin{subfigure}{.3\textwidth}
      \centering
      \includegraphics[width=.95\linewidth]{../content/cookie_figure_grid.png}  
      \caption{}
      \label{SUBFIGURE LABEL 1}
  \end{subfigure}
  \begin{subfigure}{.3\textwidth}
      \centering
      \includegraphics[width=.95\linewidth]{../content/cookie_figure_traverse_start.png}  
      \caption{}
      \label{SUBFIGURE LABEL 2}
  \end{subfigure}
  \begin{subfigure}{.3\textwidth}
      \centering
      \includegraphics[width=.95\linewidth]{../content/cookie_figure_second_row.png}  
      \caption{}
      \label{SUBFIGURE LABEL 3}
  \end{subfigure} 
  \caption{An example imaging process of a cookie sample. Here, the height and width of the capture region enclose the entire surface area of the cookie.  TIM begins by generating a grid of target coordinates to systematically capture a grid of continuous and partially overlapping images (a). TIM takes its first image at the center coordinate of the sample, then the rectangular field of view from the camera moves to adjacent target coordinates until it reaches the left boundary---capturing an image for each target coordinate (b). After completing the row, TIM moves to a new row and continues capturing adjacent images (c). }
  \label{FIGURE LABEL}
  \end{figure}

The goal of this sample transverse is to obtain in-focus images that have a region of overlap with each of its neighbors---the basis of image stitching (Figure \ref{FIGURE LABEL}). 
The coordinates of each image-to-be-captured are derived from the sample's center coordinate, the height and width of the sample, the height and width of an image frame, and the user provided percentage overlap between images.  
Our design can capture cores by defining the sample width to be zero and the height of the core to be used. So long as the cores are aligned to be parallel to the Y axis, this minimizes imaging time as only one column of images is captured.
Digitizing cookies requires a rectangular grid of images to be taken in both the X and Y axes. 
The user provides enough information to find the X and Y coordinates of each image, but the Z coordinate must be found with an algorithmic focusing procedure.

@@

With TIM's microscope lens, the depth of field of the image is sensitive enough that sub-millimeter heights can move an image out-of-focus.

This means that the height collected when adding a sample ($Z_0$, we provide a complete glossary of notation used to describe the methods, see Table \ref{tab:math_notation}) does not produce in-focus images throughout the entire surface area of the sample.
To capture an in-focus image at height $Z_{focused,k}$, for each unique image coordinate $(X_k, Y_k)$, we implemented a time-conscious software procedure. 
Our procedure involves two stages and starts at the known center coordinate of the sample.
The first stage takes advantage of the requirement that the sample is in-focus when adding it to be scanned.
Starting at the center coordinate allows the initial $Z_0$ value, captured when adding a sample, to find $Z_{focused,0}$. 
After finding an in-focus image, TIM navigates to the next neighboring image coordinate to find its in-focus image, using the previous neighbor's in-focus height as an initial approximation.

@@

The second stage of our time-efficient focusing procedure uses a feedback loop to update the initial $Z_k$ at each image location $(X_k, Y_k, Z_k)$. 
Without a feedback loop, the sample alignment must have a difference in height no greater than 0.5 mm ($d_1 - d_2 < 0.5\,\text{mm}$ in Figure \ref{fig:realistic_levelling}). 
Rather than increasing this height range, a greater alignment error can be managed by adjusting the starting image height until $Z_{focused,k}$ is at the center index of $\boldsymbol{Z_{samples,k}}$. 
The likelihood of an adjacent image, at the coordinate $(X_{k+1}, Y_{k+1})$, containing an in-focus image is highest in this scenario. 
A negative feedback  
proportional-integral-derivative (PID) control algorithm is used to adjust the initial height of each image, with $\alpha_k$, whenever the camera moves in the X or Y dimensions \citep{odwyer_summary_2000}. 
This alleviates the need of having remarkably level sample alignment ($d_1 - d_2 < 0.5\,\text{mm}$ in Figure \ref{fig:realistic_levelling}). 
\[
Z_k = \begin{cases}
  Z_0 & \text{if } k = 0,\\
  Z_{k-1} + \alpha_k \\
  \end{cases}
\]
\[\boldsymbol{\delta} = 
\begin{bmatrix}
  0.5 \\
  0.4 \\
  \vdots \\
  -0.4 \\
  -0.5 \\
  \end{bmatrix}\]
\[
Z_{focused,k} \in
\boldsymbol{Z_k} = 
Z_k + \boldsymbol{\delta}
\]

@@

After testing numerous image stitching software APIs, we found the Python package Stitch2D most capable of stitching our images successfully. % need to reference this
This package wraps OpenCV functions for finding distinctive image features using the SIFT feature detector and matches these features within the overlapping region of adjacent images \citep{lowe_distinctive_2004}. 
The default implementation of the package produces high quality stitches but is memory inefficient. %is this wording correct?, I'm trying to kindly say it was memory inefficient
We implemented a few key memory-conscious changes to the package and were able to run it on the Jetson edge computer without a problem.

@@

@@

\subsection{Strengths and Opportunities} %emw9Mar: And this section could become part of the Results & Discussion.... 
% add better opportunities
TIM's design minimizes the barriers to build an affordable, efficient and high-resolution image capture system in smaller labs without compromising scan quality, and while allowing scanning of cookies and cores. We have measured tree ring widths using the digitized samples from TIM using CooRecorder as well as ImageJ \citep{schneider_nih_2012}. 
%emw9Mar: Any citations for developing repos to share tree ring images? Or just more on building better open shared data for tree rings? 
% It appears databases store data after images are processed and measured, potentially too expensive to store image datasets at high resolution?
Publicly available tree-ring datasets can be found on multiple online repositories such as TreeSource's Tree-Rings Database for Canadian sites or the International Tree-Ring Data Bank (ITRDB) for worldwide data \citep{girardin_national_2021} \citep{grissino-mayer_international_1997}.
Both of these databases offer metadata-rich data on measured tree-rings but do not store the source images of samples.
One opportunity for this development could come as more universal and large-scale open-data storage of tree ring images come online. 
Such open repositories would also increase the value of developing new frameworks count vessel or identify rings, which are already starting and have the potential to greatly decrease the time invested in manually identifying tree anatomy \citep{resente_mask_2021} \citep{polacek_automation_2023}. 
With these more fully developed TIM could be extended to be run such frameworks while scanning, possibly taking additional scans only when needed to reach sufficient algorithm accuracy. %emw9Mar -- check my riffing here.

In its current state, TIM only provides scanned images and requires users to import them into other tools for additional visual analyses, such as measuring tree-ring widths or counting vessels.
A useful addition to TIM would be to automatically detect tree-ring widths or other metrics while capturing images using computer vision models. We expect this may be possible as these methods improve in efficiency. Currently, we found that tree-ring detection by a recent machine learning algorithm  \citeauthor[R-CNN presented by][]{polacek_automation_2023} for a scanned core (23294 x 897 pixels) took 15 minutes on a high-performance cluster (UBC ARC Sockeye) and 3.5 hours on a smaller server (RTX A6000 GPU server, which currently costs approximately \$3.50 USD for this). 
Although the Jetson edge computer is not as powerful as these servers, running these models in parallel to the image capturing algorithms would further reduce manual data annotation of tree-rings, and thus could help scale-up data collection.

@@

%emw14Apr2025: Add acknowledgments here: Deirdre, Hoai Huong Nguyen Phan, Fredi, CRD, Mao etc. 
\section{Acknowledgments}
The authors thank Deirdre Loughnan, Hoai Huong Nguyen Phan, Frederik Baumgarten for conversations and ideas to help bring this project to fruition. Additionally, we thank Xiaomao Wang and Christophe Rouleau-Desrochers for their help and patience in operating TIM during development.

\section{Supplemental Information}

{
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}  % Start supplemental figures from S1

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[height=0.5\linewidth]{../diagrams/dpi_measurement_whole.png}
    \caption{Scale Bar for DPI measurements}
    \label{fig:dpi_measurement_whole}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[height=0.5\linewidth]{../diagrams/dpi_measurement_zoomed_2.png}
    \caption{Zoomed in Scale Bar}
    \label{fig:dpi_measurement_zoomed}
  \end{subfigure}
  \caption{Example of a single DPI measurement using a 0.01 mm slide scale. We found the DPI measured at multiple locations in the field of view vertically and horizontally converged on the same value.}
  \label{fig:dpi_measurements}
\end{figure}
}

\renewcommand{\thetable}{S\arabic{table}}
\include{notation_glossary_table}
